

---
title: "Simulation study: Globally aligned PCA"
author: "Hedayat Fathi"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 3
    df_print: paged
---

This notebook reproduces the Monte Carlo simulation study in the paper
*"Globally aligned Principal Component Analysis for multi-group data."*
The core method and the performance criteria (W, PVE, A, S) live in
`Global_PCA.R` and are loaded below.

# Setup

```{r setup, message=FALSE}
library(dplyr)   # data wrangling for the result tables
library(MASS)    # mvrnorm() for data generation
library(knitr)   # kable() for table display

# Load the method and the rank-r criteria
source("Global_PCA.R")
```


# Data generation

Generate a multi-group Gaussian model with `r_true` shared global directions,
perturbed independently within each group. `sigma_theta` controls how far the
group-wise directions deviate from the shared global directions.

```{r data-generation}
generate_sim_data <- function(p            = 10,
                              K            = 8,
                              n_per_group  = rep(50, K),
                              sigma_theta  = 0.3,
                              r_true       = 3,
                              lambda_vec   = c(4, 2, 1.5),
                              lambda_noise = 1,
                              seed         = NULL) {
  if (!is.null(seed)) set.seed(seed)
  if (length(n_per_group) == 1L) n_per_group <- rep(n_per_group, K)

  # 1) Global orthonormal directions U0 (p x r_true)
  U0_raw <- matrix(rnorm(p * r_true), nrow = p, ncol = r_true)
  U0     <- qr.Q(qr(U0_raw))

  # 2) Group-specific directions: perturb then Gram-Schmidt
  u_true     <- array(NA_real_, dim = c(p, r_true, K))
  Sigma_list <- vector("list", K)

  for (k in seq_len(K)) {
    U_perturbed <- matrix(NA_real_, nrow = p, ncol = r_true)
    for (j in seq_len(r_true)) {
      eta_kj           <- rnorm(p, mean = 0, sd = sigma_theta)
      tmp              <- U0[, j] + eta_kj
      U_perturbed[, j] <- tmp / sqrt(sum(tmp^2))
    }
    # Gram-Schmidt via QR
    U_k          <- qr.Q(qr(U_perturbed))
    u_true[,, k] <- U_k

    # Sigma_k = lambda_noise * I + sum_j (lambda_j - lambda_noise) u_kj u_kj^T
    Sigma_k <- lambda_noise * diag(p)
    for (j in seq_len(r_true)) {
      u_kj    <- U_k[, j]
      Sigma_k <- Sigma_k + (lambda_vec[j] - lambda_noise) * (u_kj %*% t(u_kj))
    }
    Sigma_list[[k]] <- Sigma_k
  }

  # 3) Generate data
  X_list    <- list()
  group_vec <- character(sum(n_per_group))
  start_idx <- 1

  for (k in seq_len(K)) {
    nk      <- n_per_group[k]
    X_k     <- MASS::mvrnorm(n = nk, mu = rep(0, p), Sigma = Sigma_list[[k]])
    X_list[[k]] <- X_k
    end_idx <- start_idx + nk - 1
    group_vec[start_idx:end_idx] <- paste0("G", k)
    start_idx <- end_idx + 1
  }

  X_all  <- do.call(rbind, X_list)
  groups <- factor(group_vec, levels = paste0("G", seq_len(K)))

  list(X = X_all, groups = groups, U0 = U0,
       u_true = u_true, Sigma_true = Sigma_list, n_per_group = n_per_group)
}
```



# One replicate

Generate one dataset, then evaluate the three methods (global, group-wise,
aligned) for every `r` and every alignment strength `tau`. The criteria
`compute_*_r()` and `Global_PCA()` come from `Global_PCA.R`.

```{r one-replicate}
run_one_rep <- function(p            = 10,
                        K            = 8,
                        n_per_group  = rep(50, K),
                        sigma_theta  = 0.3,
                        tau_grid     = c(0, 0.5, 1, 2, 3),
                        r_values     = c(1, 2, 3),
                        r_true       = 3,
                        lambda_vec   = c(4, 2, 1.5),
                        lambda_noise = 1,
                        seed         = NULL) {

  sim    <- generate_sim_data(p = p, K = K, n_per_group = n_per_group,
                              sigma_theta = sigma_theta, r_true = r_true,
                              lambda_vec = lambda_vec, lambda_noise = lambda_noise,
                              seed = seed)
  X      <- sim$X; groups <- sim$groups; u_true <- sim$u_true
  K      <- length(levels(groups))

  X_scaled <- scale(X, center = TRUE, scale = FALSE)

  # Pooled (global) covariance and eigendecomposition
  Sigma_glob_hat <- cov(X_scaled)
  eig_glob_hat   <- eigen(Sigma_glob_hat, symmetric = TRUE)
  lambda_global  <- eig_glob_hat$values
  V_global_all   <- eig_glob_hat$vectors

  group_levels     <- levels(groups)
  p_dim            <- ncol(X_scaled)
  Sigma_k_hat_list <- lapply(seq_len(K), function(k) {
    idx <- which(groups == group_levels[k])
    cov(X_scaled[idx, , drop = FALSE])
  })

  results_list <- list()

  for (r in r_values) {

    V_global_r <- V_global_all[, 1:r, drop = FALSE]

    # --- Global method: same r global directions in every group ---
    V_global_list <- lapply(seq_len(K), function(k) V_global_r)

    results_list[[length(results_list) + 1L]] <- data.frame(
      method      = "global",
      r           = r,
      tau         = -1, # placeholder so this row aggregates cleanly
      mean_angle  = compute_angle_error_r(V_global_list, u_true) * 180 / pi,
      W           = compute_W_r(V_global_list, Sigma_k_hat_list),
      PVE_pct     = 100 * compute_PVE_r(V_global_list, Sigma_k_hat_list),
      A           = compute_A_r(V_global_list, V_global_r),
      S           = compute_S_r(V_global_list)
    )

    # --- Group-wise (tau = 0) and aligned (tau > 0) ---
    for (tau_val in tau_grid) {

      if (tau_val == 0) {
        # Group-wise: leading r eigenvectors of each Sigma_k_hat
        V_list <- lapply(seq_len(K), function(k) {
          eig_k <- eigen(Sigma_k_hat_list[[k]], symmetric = TRUE)
          eig_k$vectors[, 1:r, drop = FALSE]
        })
        method_name <- "groupwise"
      } else {
        # Aligned PCA with rho_m = tau * lambda_global_m
        rho_vec <- tau_val * lambda_global[1:r]
        fit_pen <- Global_PCA(X = X_scaled, groups = groups, r = r,
                                 rho = rho_vec, center = FALSE, scale. = FALSE)
        V_list <- lapply(seq_len(K), function(k) {
          V_k_raw <- fit_pen$per_group[[k]]$eig$vectors[, 1:r, drop = FALSE]
          qr.Q(qr(V_k_raw))
        })
        method_name <- "aligned"
      }

      results_list[[length(results_list) + 1L]] <- data.frame(
        method     = method_name,
        r          = r,
        tau        = tau_val,
        mean_angle = compute_angle_error_r(V_list, u_true) * 180 / pi,
        W          = compute_W_r(V_list, Sigma_k_hat_list),
        PVE_pct    = 100 * compute_PVE_r(V_list, Sigma_k_hat_list),
        A          = compute_A_r(V_list, V_global_r),
        S          = compute_S_r(V_list)
      )
    }
  }

  do.call(rbind, results_list)
}
```



# Full simulation

Repeat `run_one_rep()` over `R` replications at a fixed `sigma_theta`,
recording the replicate index and settings.

```{r full-simulation}
run_simulation <- function(R            = 200,
                           p            = 10,
                           K            = 8,
                           n_per_group  = rep(50, K),
                           sigma_theta  = 0.3,
                           tau_grid     = c(0, 0.5, 1, 2, 3),
                           r_values     = c(1, 2, 3),
                           r_true       = 3,
                           lambda_vec   = c(4, 2, 1.5),
                           lambda_noise = 1,
                           base_seed    = 2025) {
  res_list <- vector("list", R)
  for (rep in seq_len(R)) {
    cat("Replicate", rep, "of", R, "\n")
    res_list[[rep]] <- run_one_rep(
      p = p, K = K, n_per_group = n_per_group,
      sigma_theta  = sigma_theta,
      tau_grid     = tau_grid,
      r_values     = r_values,
      r_true       = r_true,
      lambda_vec   = lambda_vec,
      lambda_noise = lambda_noise,
      seed         = base_seed + rep   # distinct, reproducible seed per replicate
    )
    res_list[[rep]]$rep         <- rep
    res_list[[rep]]$sigma_theta <- sigma_theta
    res_list[[rep]]$n_group     <- n_per_group[1]
  }
  do.call(rbind, res_list)
}
```



# Run the full grid

Run once per `sigma_theta`, bind everything, and aggregate over replicates.
This is the slow step (200 replicates x 4 heterogeneity levels).

```{r run-grid}
# Fixed parameters
p            <- 10
K            <- 8
n_per_grp    <- rep(50, K)
r_true       <- 3
lambda_vec   <- c(4, 2, 1.5)
lambda_noise <- 1
R_reps       <- 200
base_seed    <- 2025

# Grid as specified in the paper
tau_grid     <- c(0, 0.5, 1, 2, 3)
r_values     <- c(1, 2, 3)
sigma_grid   <- c(0.05, 0.1, 0.3, 0.5)

sim_results_list <- vector("list", length(sigma_grid))

for (s in seq_along(sigma_grid)) {
  sigma_val <- sigma_grid[s]
  cat("=== sigma_theta =", sigma_val, "===\n")

  sim_results_list[[s]] <- run_simulation(
    R            = R_reps,
    p            = p,
    K            = K,
    n_per_group  = n_per_grp,
    sigma_theta  = sigma_val,
    tau_grid     = tau_grid,
    r_values     = r_values,
    r_true       = r_true,
    lambda_vec   = lambda_vec,
    lambda_noise = lambda_noise,
    base_seed    = base_seed
  )
}

# One data frame with everything, then aggregate over replicates
sim_results_all <- do.call(rbind, sim_results_list)

agg_results_all <- aggregate(
  cbind(mean_angle, W, PVE_pct, A, S) ~ method + r + tau + sigma_theta,
  data = sim_results_all,
  FUN  = mean
)

write.csv(agg_results_all, "sim_agg_results.csv", row.names = FALSE)
agg_results_all
```


# Tables in the paper (r = 3)

```{r table-1}
## Table 1: sigma_theta = 0.3, r = 3, trade-off across tau
tab1 <- agg_results_all %>%
  filter(sigma_theta == 0.3, r == 3) %>%
  arrange(method, tau)

knitr::kable(tab1, digits = 3,
  caption = "Table 1: sigma_theta = 0.3, r = 3. Trade-off across tau.")
```

```{r table-2}
## Table 2: r = 3, metrics across the sigma_theta grid (global + tau in {0, 1})
tab2 <- agg_results_all %>%
  filter(r == 3, method == "global" | tau %in% c(0, 1)) %>%
  arrange(sigma_theta, method, tau)

knitr::kable(tab2, digits = 3,
  caption = "Table 2: r = 3. Metrics across the sigma_theta grid.")
```

```{r table-3}
## Table 3: sigma_theta = 0.3, subspace A and S for r in {1, 2, 3}
tab3 <- agg_results_all %>%
  filter(sigma_theta == 0.3, method == "global" | tau %in% c(0, 1)) %>%
  dplyr::select(method, r, tau, A, S) %>%
  arrange(r, method, tau)

knitr::kable(tab3, digits = 3,
  caption = "Table 3: sigma_theta = 0.3, tau = 1. Subspace A and S for r in {1,2,3}.")
```


# Appendix tables: eigenvector-recovery angle (r = 3)

```{r table-A1}
## Appendix Table A1: angle error, sigma_theta = 0.3, r = 3, across tau
tab_angle_tau <- agg_results_all %>%
  filter(sigma_theta == 0.3, r == 3) %>%
  dplyr::select(method, tau, mean_angle) %>%
  arrange(method, tau)

knitr::kable(tab_angle_tau, digits = 3,
  caption = "Appendix Table A1: Angle error, sigma_theta = 0.3, r = 3, across tau.")
```

```{r table-A2}
## Appendix Table A2: angle error, r = 3, across the sigma_theta grid
tab_angle_sigma <- agg_results_all %>%
  filter(r == 3, method == "global" | tau %in% c(0, 1)) %>%
  dplyr::select(sigma_theta, method, tau, mean_angle) %>%
  arrange(sigma_theta, method, tau)

knitr::kable(tab_angle_sigma, digits = 3,
  caption = "Appendix Table A2: Angle error, r = 3, across the sigma_theta grid.")
```
